Minimal trap spaces of Boolean networks are maximal siphons of their Petri net encoding

### 2023-01-05 manuscripts/TCS2022/tcs2022.tex
Complementary places

The main property of this encoding is that it is completely faithful with respect to the update scheme of the original Boolean network.
For each node \(v\) of \(\mathcal{N}\), only transitions corresponding to \(v\) can change the current marking of \(p_v\) or \(\overline{p_v}\).
In addition, at any marking at most one of such transitions is enabled because \(m(p_v) + m(\overline{p}_v) = 1\) holds.
Hence, for any update scheme in \(\mathcal{N}\), we have a corresponding firing scheme in \(\mathcal{P}\), which preserves the equivalence between the dynamics of \(\mathcal{N}\) and \(\mathcal{P}\)~\cite{DBLP:journals/nc/ChatainHKPT20}.

For illustration, let us reconsider the Boolean network shown in Example~\ref{example:BN}.
Figure~\ref{fig:PN} shows the Petri net encoding of this Boolean network.
Place \(p_{x_1}\) (resp. \(\overline{p_{x_1}}\)) in \(\mathcal{P}\) represents the activation (resp.\ the inactivation) of node \(x_1\) in \(\mathcal{N}\).
Marking \(\{p_{x_1}, \overline{p_{x_2}}\}\) in \(\mathcal{P}\) represents state 10 in \(\mathcal{N}\).
Transitions \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) represent the update of node \(x_1\).
Of course, in any marking \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) cannot be both enabled.
Then, the fully asynchronous update scheme in \(\mathcal{N}\) corresponds to the classical firing scheme in \(\mathcal{P}\) where only one of the enabled transitions for a given marking will be fired~\cite{Murata1989}.

@@

Intuitively, a siphon is a set of places that once unmarked remains so.
If it is conflict-free then its dual corresponds to a partial-state of the model such that whatever update, the fixed values remain so (since the unmarked places remain unmarked).
This is precisely the definition of a trap space and maximality of the siphon is equivalent to as many fixed values as possible, hence minimality of the trap space.
For example, the Boolean network given in Example~\ref{example:BN} has two trap spaces, \(m_1 = 11\) and \(m_2 = \star\star\).
The Petri net encoding of this Boolean network has five generic siphons, \(S_1 = \emptyset\), \(S_2 = \{p_{x_1}, \overline{p_{x_1}}\}\), \(S_3 = \{p_{x_2}, \overline{p_{x_2}}\}\), \(S_4 = \{\overline{p_{x_1}}, \overline{p_{x_2}}\}\), and \(S_5 = \{p_{x_1}, \overline{p_{x_1}}, p_{x_2}, \overline{p_{x_2}}\}\).
However, only \(S_1\) and \(S_4\) are conflict-free siphons and correspond to \(m_2\) and \(m_1\), respectively.
Since \(S_1 \subset S_4\), \(S_4\) is a maximal siphon corresponding to the minimal trap space \(m_1\).
Hereafter, we formally prove that a maximal conflict-free siphon is equivalent to a minimal trap space.

@@

First, we show the characterization of all conflict-free siphons of the encoded Petri net \(\mathcal{P} = (P, T, W)\).
Suppose that \(S\) is a generic siphon of \(\mathcal{P}\). If a place \(p\) should belong to \(S\), then by Proposition~\ref{prop:siphon_set} all the transitions in \(pred(p)\) must belong to \(succ(S)\).
A transition \(t\) belongs to \(succ(S)\) if and only if there is at least one place \(p'\) in \(S\) such that \(p' \in pred(t)\).
Hence, for each transition \(t \in pred(p)\), we can state that
\begin{equation}
\label{eq:siphon}
p \in S \Rightarrow \bigvee_{p' \in pred(t)}p' \in S.
\end{equation}
The system of all the rules of the above form with respect to all pairs \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\) fully characterizes all generic siphons of a Petri net and has been used with SAT solvers in~\cite{oanea2010new,nabli2016enumerating}.
To make \(S\) to be a conflict-free siphon, we need to add to the system the rule
\begin{equation}
\label{eq:conflict}
p_v \in S \Rightarrow \overline{p_v} \not \in S \wedge \overline{p_v} \in S \Rightarrow p_v \not \in S
\end{equation}for each node \(v \in V\).
By definition, the final system fully characterizes all conflict-free siphons of the encoded Petri net.

@@

  Given a Petri net \(\mathcal{P} = (P, T, W)\) encoding a Boolean network \(\mathcal{N} = (V, F)\).
  The CSP \(\mathcal{C}(\mathcal{P})\) is the triple \((R, D, C)\) where
  \begin{itemize}%
    \item \(R = P\), i.e., a variable is introduced for each place of \(\mathcal{P}\),
    \item \(D(p) = \mathbb{B}\) for all \(p \in R\), i.e., the variables are Boolean,
    \item \(C = \{\neg p_v \vee \neg \overline{p_v} = 1 \mid \forall v \in V\} \bigwedge
\{(p = 1 \rightarrow \bigvee_{p' \in pred(t)}p' = 1) \mid p \in P, t \in pred(p)\}\).
  \end{itemize}

@@

  By the former part \(\neg p_v \vee \neg \overline{p_v}\) = 1 of \(C\), the conflict-freeness is imposed because for any satisfable valuation \(r\), \(r(p_v) = r(\overline{p_v}) = 1\) is impossible for all \(v \in V\).
  As shown in~\cite{nabli2016enumerating}, the latter part of \(C\) can characterize the set of all generic siphons of \(\mathcal{P}\).
  Hence, we can conclude the proof.

@@

One natural question is that how to use the CSP-based method for enumerating all the maximal conflict-free siphons of a Petri net encoding a Boolean model?
Of course, the set of all conflict-free siphons of the Petri net can easily characterized by the CSP model presented in~\cite{nabli2016enumerating} along with the additional constraint \(\neg p_v \vee \neg \overline{p_v} = 1\), for each \(v \in V\), which represents the conflict-freeness.
However, the main concern is to enumerate all the \emph{maximal} ones, which is not trivial to adapt from the CSP-based method.
By Proposition~\ref{prop:csp_conflict_free_siphon}, the set of all maximal conflict-free siphons of \(\mathcal{P}\) can be enumerated in the (maximality) set inclusion order, by restarting the search each time a conflict-free siphon \(S\) is found, with the following additional constraint for disallowing any subset of that conflict-free siphon: \(\bigvee_{p \not \in S} p = 1\).
For enumerating conflict-free siphons in the set inclusion order, we can use the same technique as used in~\cite{nabli2016enumerating} but with the opposite setting, i.e., labeling directly the Boolean variables with decreasing value selection.
The correctness of this technique comes from the fact that once \(S\) is found, it is the conflict-free siphon of maximum cardinality among all the remaining feasible conflict-free siphons.
Similar to~\cite{nabli2016enumerating}, the newly CSP-based method can also be implemented with SAT and CLP solvers.

However a more direct method is to use a MaxSAT solver.
We construct a MaxSAT problem with the following hard clauses:
\[
  (\neg p_v \vee \neg \overline{p_v}), \forall v \in V
\]
and
\[
  (\neg p \vee \bigvee_{p' \in pred(t)}p'), \forall p \in P, \forall t \in pred(p).
\]
We set a soft clause for each variable of the CSP and then use a ``minimal correction subset'' blocking strategy, which will ensure set-inclusion maximality of the solutions.
This is what is implemented in \texttt{trappist} using the RC2 MaxSAT solver~\cite{DBLP:journals/jsat/IgnatievMM19} available through the \texttt{python-sat} package\footnote{\url{https://pysathq.github.io/docs/html/api/examples/rc2.html}}.

@@

Another possible method is to translate the characterization shown in Subsection~\ref{subsec:siphon_characterization} into the ASP \(\mathcal{L}\) as follows.
We introduce atom \verb|p-v| (resp. \verb|n-v|) to denote place \(p_v\) (resp. \(\overline{p_v}\)), \(\forall v \in V\). 
The set of all atoms in \(\mathcal{L}\) is given as \(\mathcal{A} = \bigcup_{v \in V}\{\verb|p-v|, \verb|n-v|\}\).
For each pair \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\), we translate the rule (\ref{eq:siphon}) into the ASP rule
\[
\verb|a_1; ... ; a_k :- a.|
\]
where \(\verb|a| \in \mathcal{A}\) is the atom representing place \(p\) and \(\{\verb|a_1|, \dots,\verb|a_k|\} \subseteq \mathcal{A}\) is the set of atoms representing places in \(pred(t)\). The rule (\ref{eq:conflict}) is translated into the ASP rule
\[\verb|:- p-v, n-v.|\]
for each \(v \in V\). 
This ASP rule guarantees that two places representing the same node in \(\mathcal{N}\) never belong to the same siphon of \(\mathcal{P}\), representing the conflict-freeness.
Naturally, a Herbrand model (see, e.g.,~\cite{DBLP:journals/aicom/GebserKKOSS11}) of \(\mathcal{L}\) is equivalent to a conflict-free siphon of \(\mathcal{P}\).
To guarantee that a Herbrand model is also a stable model (an answer set), we need to add to \(\mathcal{L}\) the two choice rules
\[
\verb|{p-v}. {n-v}.|
\]
for each \(v \in V\).
Note that the number of atoms of \(\mathcal{L}\) is only \(2n\), whereas the ASP encoding shown in~\cite{klarner2015computing} has as many atoms as the number of prime-implicants of the Boolean network and that number might be exponential in \(n\).
In~\cite{DBLP:conf/ictai/ChevalierFPZ19}, there is an ASP characterization of trap spaces that does not rely on minimal DNFs either and thus seems very similar to our ASP encoding.
Remarkably it only requires the DNF for the \emph{activation} part, using the information that it will only be used for locally-monotonic Boolean networks.
We would therefore expect that, when available, it will have comparable performance on the ASP part (the ASP program would be approximately twice smaller, though redundancy is not always bad in that field), but can also avoid combinatorial explosion of the Petri net encoding for some formula where the activation DNF is simple but the inhibition is not.
Since mpbn is included in our benchmark this will be evaluated in our experiments.

@@

We first show how an Integer Linear Programming (ILP) \(\mathcal{I}\) can define a set of all conflict-free siphons of the encoded Petri net \(\mathcal{P}\).
We introduce \emph{binary} variable \verb|p-v| (resp. \verb|n-v|) to denote place \(p_v\) (resp. \(\overline{p_v}\)), \(\forall v \in V\).
The set of all binary variables in \(\mathcal{I}\) is \(\bigcup_{v \in V}\{\verb|p-v|, \verb|n-v|\}\).
For each pair \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\), we translate the rule (\ref{eq:siphon}) into the ILP inequality
\[
\verb|a <= a_1 + ... + a_k|
\]
where \verb|a| is the binary variable representing place \(p\) and \{\verb|a_1|, \dots,\verb|a_k|\} is the set of binary variable representing places in \(pred(t)\).
The rule (\ref{eq:conflict}) is translated into the ILP inequality
\[\verb|p-v + n-v <= 1|\]
for each \(v \in V\).
This inequality forbids both \verb|p-v| and \verb|n-p| receive the value 1, thus representing the conflict-freeness.
Since we only consider feasible solutions, the objective function is set to \verb|max p-v| for some \(v \in V\).
Naturally, a solution \(I\) of \(\mathcal{I}\) is equivalent to a conflict-free siphon \(S\) of \(\mathcal{P}\).
The conversion is that
\[
  S = \{p \in P \;|\; I(\verb|a-p|) = 1\}
\]
where \verb|a-p| is the binary variable presenting place \(p\).

>>>>

The main property of this encoding is that it is completely faithful with respect to the update scheme of the original Boolean network.
For each node \(v\) of \(\mathcal{N}\), only transitions corresponding to \(v\) can change the current marking of \(p_v\) or \(\overline{p}_v\).
In addition, at any marking at most one of such transitions is enabled because \(m(p_v) + m(\overline{p}_v) = 1\) holds.
Hence, for any update scheme in \(\mathcal{N}\), we have a corresponding firing scheme in \(\mathcal{P}\), which preserves the equivalence between the dynamics of \(\mathcal{N}\) and \(\mathcal{P}\)~\cite{DBLP:journals/nc/ChatainHKPT20}.

For illustration, let us reconsider the Boolean network shown in Example~\ref{example:BN}.
Figure~\ref{fig:PN} shows the Petri net encoding of this Boolean network.
Place \(p_{x_1}\) (resp. \(\overline{p}_{x_1}\)) in \(\mathcal{P}\) represents the activation (resp.\ the inactivation) of node \(x_1\) in \(\mathcal{N}\).
Marking \(\{p_{x_1}, \overline{p}_{x_2}\}\) in \(\mathcal{P}\) represents state 10 in \(\mathcal{N}\).
Transitions \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) represent the update of node \(x_1\).
Of course, in any marking \(t^{1}_{x_1}\) and \(t^{2}_{x_1}\) cannot be both enabled.
Then, the fully asynchronous update scheme in \(\mathcal{N}\) corresponds to the classical firing scheme in \(\mathcal{P}\) where only one of the enabled transitions for a given marking will be fired~\cite{Murata1989}.

@@

Intuitively, a siphon is a set of places that once unmarked remains so.
If it is conflict-free then its dual corresponds to a partial-state of the model such that whatever update, the fixed values remain so (since the unmarked places remain unmarked).
This is precisely the definition of a trap space and maximality of the siphon is equivalent to as many fixed values as possible, hence minimality of the trap space.
For example, the Boolean network given in Example~\ref{example:BN} has two trap spaces, \(m_1 = 11\) and \(m_2 = \star\star\).
The Petri net encoding of this Boolean network has five generic siphons, \(S_1 = \emptyset\), \(S_2 = \{p_{x_1}, \overline{p}_{x_1}\}\), \(S_3 = \{p_{x_2}, \overline{p}_{x_2}\}\), \(S_4 = \{\overline{p}_{x_1}, \overline{p}_{x_2}\}\), and \(S_5 = \{p_{x_1}, \overline{p}_{x_1}, p_{x_2}, \overline{p}_{x_2}\}\).
However, only \(S_1\) and \(S_4\) are conflict-free siphons and correspond to \(m_2\) and \(m_1\), respectively.
Since \(S_1 \subset S_4\), \(S_4\) is a maximal siphon corresponding to the minimal trap space \(m_1\).
Hereafter, we formally prove that a maximal conflict-free siphon is equivalent to a minimal trap space.

@@

First, we show the characterization of all conflict-free siphons of the encoded Petri net \(\mathcal{P} = (P, T, W)\).
Suppose that \(S\) is a generic siphon of \(\mathcal{P}\). If a place \(p\) should belong to \(S\), then by Proposition~\ref{prop:siphon_set} all the transitions in \(pred(p)\) must belong to \(succ(S)\).
A transition \(t\) belongs to \(succ(S)\) if and only if there is at least one place \(p'\) in \(S\) such that \(p' \in pred(t)\).
Hence, for each transition \(t \in pred(p)\), we can state that
\begin{equation}
\label{eq:siphon}
p \in S \Rightarrow \bigvee_{p' \in pred(t)}p' \in S.
\end{equation}
The system of all the rules of the above form with respect to all pairs \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\) fully characterizes all generic siphons of a Petri net and has been used with SAT solvers in~\cite{oanea2010new,nabli2016enumerating}.
To make \(S\) to be a conflict-free siphon, we need to add to the system the rule
\begin{equation}
\label{eq:conflict}
p_v \in S \Rightarrow \overline{p}_v \not \in S \wedge \overline{p}_v \in S \Rightarrow p_v \not \in S
\end{equation}for each node \(v \in V\).
By definition, the final system fully characterizes all conflict-free siphons of the encoded Petri net.

@@

  Given a Petri net \(\mathcal{P} = (P, T, W)\) encoding a Boolean network \(\mathcal{N} = (V, F)\).
  The CSP \(\mathcal{C}(\mathcal{P})\) is the triple \((R, D, C)\) where
  \begin{itemize}%
    \item \(R = P\), i.e., a variable is introduced for each place of \(\mathcal{P}\),
    \item \(D(p) = \mathbb{B}\) for all \(p \in R\), i.e., the variables are Boolean,
    \item \(C = \{\neg p_v \vee \neg \overline{p}_v = 1 \mid \forall v \in V\} \bigwedge
\{(p = 1 \rightarrow \bigvee_{p' \in pred(t)}p' = 1) \mid p \in P, t \in pred(p)\}\).
  \end{itemize}

@@

  By the former part \(\neg p_v \vee \neg \overline{p}_v\) = 1 of \(C\), the conflict-freeness is imposed because for any satisfable valuation \(r\), \(r(p_v) = r(\overline{p}_v) = 1\) is impossible for all \(v \in V\).
  As shown in~\cite{nabli2016enumerating}, the latter part of \(C\) can characterize the set of all generic siphons of \(\mathcal{P}\).
  Hence, we can conclude the proof.

@@

One natural question is that how to use the CSP-based method for enumerating all the maximal conflict-free siphons of a Petri net encoding a Boolean model?
Of course, the set of all conflict-free siphons of the Petri net can easily characterized by the CSP model presented in~\cite{nabli2016enumerating} along with the additional constraint \(\neg p_v \vee \neg \overline{p}_v = 1\), for each \(v \in V\), which represents the conflict-freeness.
However, the main concern is to enumerate all the \emph{maximal} ones, which is not trivial to adapt from the CSP-based method.
By Proposition~\ref{prop:csp_conflict_free_siphon}, the set of all maximal conflict-free siphons of \(\mathcal{P}\) can be enumerated in the (maximality) set inclusion order, by restarting the search each time a conflict-free siphon \(S\) is found, with the following additional constraint for disallowing any subset of that conflict-free siphon: \(\bigvee_{p \not \in S} p = 1\).
For enumerating conflict-free siphons in the set inclusion order, we can use the same technique as used in~\cite{nabli2016enumerating} but with the opposite setting, i.e., labeling directly the Boolean variables with decreasing value selection.
The correctness of this technique comes from the fact that once \(S\) is found, it is the conflict-free siphon of maximum cardinality among all the remaining feasible conflict-free siphons.
Similar to~\cite{nabli2016enumerating}, the newly CSP-based method can also be implemented with SAT and CLP solvers.

However a more direct method is to use a MaxSAT solver.
We construct a MaxSAT problem with the following hard clauses:
\[
  (\neg p_v \vee \neg \overline{p}_v), \forall v \in V
\]
and
\[
  (\neg p \vee \bigvee_{p' \in pred(t)}p'), \forall p \in P, \forall t \in pred(p).
\]
We set a soft clause for each variable of the CSP and then use a ``minimal correction subset'' blocking strategy, which will ensure set-inclusion maximality of the solutions.
This is what is implemented in \texttt{trappist} using the RC2 MaxSAT solver~\cite{DBLP:journals/jsat/IgnatievMM19} available through the \texttt{python-sat} package\footnote{\url{https://pysathq.github.io/docs/html/api/examples/rc2.html}}.

@@

Another possible method is to translate the characterization shown in Subsection~\ref{subsec:siphon_characterization} into the ASP \(\mathcal{L}\) as follows.
We introduce atom \verb|p-v| (resp. \verb|n-v|) to denote place \(p_v\) (resp. \(\overline{p}_v\)), \(\forall v \in V\). 
The set of all atoms in \(\mathcal{L}\) is given as \(\mathcal{A} = \bigcup_{v \in V}\{\verb|p-v|, \verb|n-v|\}\).
For each pair \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\), we translate the rule (\ref{eq:siphon}) into the ASP rule
\[
\verb|a_1; ... ; a_k :- a.|
\]
where \(\verb|a| \in \mathcal{A}\) is the atom representing place \(p\) and \(\{\verb|a_1|, \dots,\verb|a_k|\} \subseteq \mathcal{A}\) is the set of atoms representing places in \(pred(t)\). The rule (\ref{eq:conflict}) is translated into the ASP rule
\[\verb|:- p-v, n-v.|\]
for each \(v \in V\). 
This ASP rule guarantees that two places representing the same node in \(\mathcal{N}\) never belong to the same siphon of \(\mathcal{P}\), representing the conflict-freeness.
Naturally, a Herbrand model (see, e.g.,~\cite{DBLP:journals/aicom/GebserKKOSS11}) of \(\mathcal{L}\) is equivalent to a conflict-free siphon of \(\mathcal{P}\).
To guarantee that a Herbrand model is also a stable model (an answer set), we need to add to \(\mathcal{L}\) the two choice rules
\[
\verb|{p-v}. {n-v}.|
\]
for each \(v \in V\).
Note that the number of atoms of \(\mathcal{L}\) is only \(2n\), whereas the ASP encoding shown in~\cite{klarner2015computing} has as many atoms as the number of prime-implicants of the Boolean network and that number might be exponential in \(n\).
In~\cite{DBLP:conf/ictai/ChevalierFPZ19}, there is an ASP characterization of trap spaces that does not rely on minimal DNFs either and thus seems very similar to our ASP encoding.
Remarkably it only requires the DNF for the \emph{activation} part, using the information that it will only be used for locally-monotonic Boolean networks.
We would therefore expect that, when available, it will have comparable performance on the ASP part (the ASP program would be approximately twice smaller, though redundancy is not always bad in that field), but can also avoid combinatorial explosion of the Petri net encoding for some formula where the activation DNF is simple but the inhibition is not.
Since mpbn is included in our benchmark this will be evaluated in our experiments.

@@

We first show how an Integer Linear Programming (ILP) \(\mathcal{I}\) can define a set of all conflict-free siphons of the encoded Petri net \(\mathcal{P}\).
We introduce \emph{binary} variable \verb|p-v| (resp. \verb|n-v|) to denote place \(p_v\) (resp. \(\overline{p}_v\)), \(\forall v \in V\).
The set of all binary variables in \(\mathcal{I}\) is \(\bigcup_{v \in V}\{\verb|p-v|, \verb|n-v|\}\).
For each pair \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\), we translate the rule (\ref{eq:siphon}) into the ILP inequality
\[
\verb|a <= a_1 + ... + a_k|
\]
where \verb|a| is the binary variable representing place \(p\) and \{\verb|a_1|, \dots,\verb|a_k|\} is the set of binary variable representing places in \(pred(t)\).
The rule (\ref{eq:conflict}) is translated into the ILP inequality
\[\verb|p-v + n-v <= 1|\]
for each \(v \in V\).
This inequality forbids both \verb|p-v| and \verb|n-p| receive the value 1, thus representing the conflict-freeness.
Since we only consider feasible solutions, the objective function is set to \verb|max p-v| for some \(v \in V\).
Naturally, a solution \(I\) of \(\mathcal{I}\) is equivalent to a conflict-free siphon \(S\) of \(\mathcal{P}\).
The conversion is that
\[
  S = \{p \in P \;|\; I(\verb|a-p|) = 1\}
\]
where \verb|a-p| is the binary variable presenting place \(p\).
